Supporting Information for "Geometric models overestimate lake depth due to imperfect slope measurement"

### 2020-11-28 manuscript/agujournaltemplate.tex
text edits

We used our generated bathymetry surfaces to find the location of the deepest point in the lake and we resolved ties by choosing the deepest point that was closest to the center of the lake.
To account for lakes where the centroid does not intersect lake bathymetry because it is located within an embedded island or peninsula, we calculated the center of the lake not as its centroid but rather by finding the point farthest from the lake shoreline (i.e. its “visual distance to lake center”). For these calculations, we used the \texttt{polylabelr} R package \cite{larssonPolylabelrFindPole2019}, which interfaces with the Mapbox pole of inaccessibility algorithm \cite{agafonkinJSLibraryFinding2019}. We calculated in-lake slope as maximum lake depth divided by the distance to the deepest point where maximum lake depth is a point measurement derived from a bathymetric surface and distance to the deepest point is the smallest straight-line distance from that point to the waterbody perimeter. We calculated nearshore land slope for each lake by computing the slope within a 100-m buffer using data from a high resolution digital elevation model ($\sim$15x15m grain) accessed using the \texttt{elevatr} R package \cite{hollisterElevatrAccessElevation2017}. Slope computations proceeded by passing a 3x3 moving window over the 100-m buffer to calculate the slope at each point using Horn's algorithm via the \texttt{terrain} function in the \texttt{raster} R package \cite{hijmansRasterGeographicData2019}. Reported nearshore land slope values are the mean of all points in the buffer. In addition to the aforementioned techniques of calculating in-lake (and nearshore) slopes and distances, we tried 7 alternate techniques which are described in Table S2.

@@

We further classified lakes using the output of a deep convolutional neural network model trained on satellite images labeled according to whether there was visual evidence of a water control structure significantly impacting flow. This model had an overall validation accuracy of 81\% and produced a probability for each waterbody as to whether it is a reservoir or a natural lake. For our purposes, we set a conservative classification probability threshold of 0.75 to determine whether a lake would be considered a  reservoir. We tested the sensitivity of our model output relative to variations of this threshold and found little substantive difference in results. Note that our reservoir classification defines reservoirs as any permanent waterbody that has a water control structure likely to significantly impact flow or pool water. It makes no distinction between different dam types, heights, or uses/purposes.

@@

\subsection{Model description}
\subsubsection{Geometric model}
\noindent
We used a geometric model of lakes where basins are treated as cones with a fixed relationship between slope and distance (Equation \ref{eq1}). One reason that we used the cone model is that, unlike other idealized shapes, it is well suited for modeling maximum depth and does not require any knowledge of lake volume or mean depth. Note that Equation \ref{eq1} is a geometric formula and has no intercept or “coefficients” and it produces a perfect depth estimate given true values of slope and distance. To use this model to predict the depth of all lakes, there is a necessary assumption that proxy slope and distance measures, which are available for all lakes, are representative of true slope and distance (Figure \ref{fig1}).

@@

In addition to overall differences between slope and distance measures, we found differences in these relationships among lake shape classes. For example, in-lake slope and distance to the deepest point of the lake metrics were consistently larger in magnitude for convex lakes as compared to concave lakes (Figure S4). However,  there were not similar differences among slope and distance metrics for natural lakes versus reservoirs (Figure S4).

@@

The most important covariates in offset models were those relating to spatial location, lake area, and perimeter (Figure S5). Conversely, watershed metrics and lake elevation had little contribution to random forest model fit (Figure S5). The spatial location (i.e. HUC4) covariate was notably less importance in the true slope model compared to the two proxy slope models. Model importance calculations indicated that omitting a geometric max depth measure results in a 130\%, 60\%, or 50\% increase in mean square error depending on the formulation of geometric max depth in Equation \ref{eq1} (Figure S5).

\section{Discussion}
\noindent
Our tests of geometric lake depth models show that specific proxy measures of lake geometry are not representative of true geometry measures across a broad array of lakes. Models using non-representative proxies showed increased error and systematic overestimation of depth in concave and reservoir lakes. Although our analysis was limited to lakes with available bathymetry data, these lakes did not have characteristics that differed from that of the overall lake population (Figure S6, S8). Although there is a possibility that there is some hidden bias not explored for in our analyses, this lack of difference suggests that our results are likely to be broadly applicable to all lakes.

\subsection{Representativeness of proxy measures of lake geometry}
\noindent
In comparing among lake geometry measures, our analysis suggests that proxy distance to lake center is representative of true distance to the deepest point of the lakes but that proxy nearshore land slope is not representative of true in-lake slope. A simple indication of this non representativeness is that proxy nearshore land slope was often (in $>$ 74\% of cases) steeper than true in-lake slope. This finding is consistent with \citeA{heathcotePredictingBathymetricFeatures2015} whos results suggest that in-lake slopes are shallower compared to the surrounding land. Furthermore, the fact that in-lake slopes were shallower compared to the surrounding land even after controlling for differences in area and depth (Figure S9) is consistent with the idea of topographic scaling (i.e. scale invariance) explored in previous work and detailed by \cite{caelVolumeMeanDepth2017}. The underlying reason for these shallow in-lake slopes may be related to slope-induced turbidity currents which distribute sediment from shallow high-energy areas of lakes to deep low-energy areas \cite{hakansonLakeBottomDynamics1981, johanssonNewApproachesModelling2007}. The strength of such sediment focusing is likely greater in "younger" lakes with steeper slopes leading to a smoothing of their bathymetry over time \cite{blaisInfluenceLakeMorphometry1995}.

@@

We found other differences among lake geometry measures according to lake cross-section shape. One finding was that convex lakes, when compared to concave lakes, had longer distances to lake centers relative to corresponding distances to the deepest point of lakes. In addition, convex lakes often had steeper in-lake slopes relative to nearshore land slopes as compared to concave lakes. Finally, it was notable that convex lakes were deeper than concave lakes despite having similar distributions of lake surface area (Figure S7). The underlying cause of these differences is unknown but one possibility is that geometry is tied to the circumstances of lake formation whereby the formation of concave lakes were a result of more intense glacial scouring compared to that of convex lakes \cite{gorhamPhysicalLimnologyNorthern1958}. While our findings provide some evidence in support of this idea, namely that there is a geographic hotspot of concave lakes associated with the glaciated “prairie pothole region” \cite<see>{hayashiSimpleEquationsRepresent2000}, the overall geographic distribution of lake cross-section shapes does not support this idea. Instead of a concentrated area of concave lakes in formerly glaciated regions, there appears to be a fairly even mix of concave and convex lakes distributed amongst the northern (i.e. glaciated) and southern (non-glaciated) portions of our study area (Figure S2).

\subsection{Effects of proxy measures of lake geometry depth prediction error}
\noindent
Models using only proxy variables had prediction error rates (RMSE = 6.4m) of a similar magnitude as that of prior studies (RMSE = 6 - 7.3m) predicting lake depth at broad geographic extents \cite{hollisterPredictingMaximumLake2011, oliver2016prediction, messagerEstimatingVolumeAge2016}. When only a single proxy measure was used, there was a difference in model sensitivity depending on if it was a horizontal distance measure or a vertical slope measure. In the case of a true slope and proxy distance combination, models were more accurate ($\pm$ 4.4m, 29\%) than even the most accurate of prior studies \cite{hollisterPredictingMaximumLake2011, oliver2016prediction, messagerEstimatingVolumeAge2016}. Conversely, models using a proxy slope and true distance combination had prediction error rates ($\pm$ 6.6m, 60\%) of a similar magnitude as that of the baseline proxy-proxy model ($\pm$ 6.4m, 59\%). The greater sensitivity of depth predictions to proxy slope measures relative to proxy distance measures may be explained by the fact that proxy slope measures were a more imperfect representation of true in-lake slopes relative to proxy versus true distances. In addition, these results help explain the relatively poor predictive performance of prior non-geometric lake depth models given that they rely heavily on lake area as a predictor \cite{messagerEstimatingVolumeAge2016, oliver2016prediction, sobekPredictingDepthVolume2011} and both horizontal distance measures and vertical slope measures appear to be decoupled from lake area (Figure S7).

>>>>

We used our generated bathymetry surfaces to find the location of the deepest point in the lake and we resolved ties by choosing the deepest point that was closest to the center of the lake. We used the location of this deepest point to calculate "distance to the deepest point" as the minimum distance to the lake shoreline. To account for lakes where the centroid does not intersect lake bathymetry because it is located within an embedded island or peninsula, we calculated the center of the lake not as its centroid but rather by finding the point farthest from the lake shoreline (i.e. its “visual distance to lake center”). For these calculations, we used the \texttt{polylabelr} R package \cite{larssonPolylabelrFindPole2019}, which interfaces with the Mapbox pole of inaccessibility algorithm \cite{agafonkinJSLibraryFinding2019}. We calculated in-lake slope as maximum lake depth divided by the distance to the deepest point where maximum lake depth is a point measurement derived from a bathymetric surface and distance to the deepest point is the smallest straight-line distance from that point to the waterbody perimeter. We calculated nearshore land slope for each lake by computing the slope within a 100-m buffer using data from a high resolution digital elevation model ($\sim$15x15m grain) accessed using the \texttt{elevatr} R package \cite{hollisterElevatrAccessElevation2017}. Slope computations proceeded by passing a 3x3 moving window over the 100-m buffer to calculate the slope at each point using Horn's algorithm via the \texttt{terrain} function in the \texttt{raster} R package \cite{hijmansRasterGeographicData2019}. Reported nearshore land slope values are the mean of all points in the buffer. In addition to the aforementioned techniques of calculating in-lake (and nearshore) slopes and distances, we tried 7 alternate techniques which are described in Table S2.

@@

We further classified lakes using the output of a deep convolutional neural network model trained on satellite images labeled according to whether there was visual evidence of a water control structure significantly impacting flow. This model had an overall validation accuracy of 81\% and produced a probability for each waterbody as to whether it is a reservoir or a natural lake. For our purposes, we set a conservative classification probability threshold of 0.75 to determine whether a lake would be considered a  reservoir. Note that our reservoir classification defines reservoirs as any permanent waterbody that has a water control structure likely to significantly impact flow or pool water. It makes no distinction between different dam types, heights, or uses/purposes.
% We tested the sensitivity of our model output relative to variations of this threshold and found little substantive difference in results.

@@

\subsection{Model description}
\subsubsection{Geometric model}
\noindent
We used a geometric model of lakes where basins are treated as cones with a fixed relationship between slope and distance (Equation \ref{eq1}). One reason that we used the cone model is that, unlike other idealized shapes, it is well suited for modeling maximum depth and does not require any knowledge of lake volume or mean depth. Note that Equation \ref{eq1} is a geometric formula and has no intercept or coefficients and it produces a perfect depth estimate given true values of slope and distance. To use this model to predict the depth of all lakes, there is a necessary assumption that proxy slope and distance measures, which are available for all lakes, are representative of true slope and distance (Figure \ref{fig1}).

@@

In addition to overall differences between slope and distance measures, we found differences in these relationships among lake shape classes. For example, in-lake slope and distance to the deepest point of the lake metrics were consistently larger in magnitude for convex lakes as compared to concave lakes (Figure S4). We found evidence that this difference was at leat partly explained by the fact that convex lakes are deeper than concave lakes (Figure S5). There were not similar differences among slope and distance metrics for natural lakes versus reservoirs.

@@

The most important covariates in offset models were those relating to spatial location, lake area, and perimeter (Figure S7). Conversely, watershed metrics and lake elevation had little contribution to random forest model fit (Figure S7). The spatial location (i.e. HUC4) covariate was notably less importance in the true slope model compared to the two proxy slope models. Model importance calculations indicated that omitting a geometric max depth measure results in a 130\%, 60\%, or 50\% increase in mean square error depending on the formulation of geometric max depth in Equation \ref{eq1} (Figure S7).

\section{Discussion}
\noindent
Our tests of geometric lake depth models show that specific proxy measures of lake geometry are not representative of true geometry measures across a broad array of lakes. Models using non-representative proxies showed increased error and systematic overestimation of depth in concave and reservoir lakes. Although our analysis was limited to lakes with available bathymetry data, these lakes did not have characteristics that differed from that of the overall lake population (Figure S8, S10). Although there is a possibility that there is some hidden bias not explored for in our analyses, this lack of difference suggests that our results are likely to be broadly applicable to all lakes.

\subsection{Representativeness of proxy measures of lake geometry}
\noindent
In comparing among lake geometry measures, our analysis suggests that proxy distance to lake center is representative of true distance to the deepest point of the lakes but that proxy nearshore land slope is not representative of true in-lake slope. A simple indication of this non representativeness is that proxy nearshore land slope was often (in $>$ 74\% of cases) steeper than true in-lake slope. This finding is consistent with \citeA{heathcotePredictingBathymetricFeatures2015} whos results suggest that in-lake slopes are shallower compared to the surrounding land. Furthermore, the fact that in-lake slopes were shallower compared to the surrounding land even after controlling for differences in area (Figure S5) is consistent with the idea of topographic scaling (i.e. scale invariance) explored in previous work and detailed by \cite{caelVolumeMeanDepth2017}. The underlying reason for these shallow in-lake slopes may be related to slope-induced turbidity currents which distribute sediment from shallow high-energy areas of lakes to deep low-energy areas \cite{hakansonLakeBottomDynamics1981, johanssonNewApproachesModelling2007}. The strength of such sediment focusing is likely greater in "younger" lakes with steeper slopes leading to a smoothing of their bathymetry over time \cite{blaisInfluenceLakeMorphometry1995}.

@@

We found other differences among lake geometry measures according to lake cross-section shape. One finding was that convex lakes, when compared to concave lakes, had longer distances to lake centers relative to corresponding distances to the deepest point of lakes. In addition, convex lakes often had steeper in-lake slopes relative to nearshore land slopes as compared to concave lakes. Finally, it was notable that convex lakes were deeper than concave lakes despite having similar distributions of lake surface area (Figure S9). The underlying cause of these differences is unknown but one possibility is that geometry is tied to the circumstances of lake formation whereby the formation of concave lakes were a result of more intense glacial scouring compared to that of convex lakes \cite{gorhamPhysicalLimnologyNorthern1958}. While our findings provide some evidence in support of this idea, namely that there is a geographic hotspot of concave lakes associated with the glaciated “prairie pothole region” \cite<see>{hayashiSimpleEquationsRepresent2000}, the overall geographic distribution of lake cross-section shapes does not support this idea. Instead of a concentrated area of concave lakes in formerly glaciated regions, there appears to be a fairly even mix of concave and convex lakes distributed amongst the northern (i.e. glaciated) and southern (non-glaciated) portions of our study area (Figure S2).

\subsection{Effects of proxy measures of lake geometry depth prediction error}
\noindent
Models using only proxy variables had prediction error rates (RMSE = 6.4m) of a similar magnitude as that of prior studies (RMSE = 6 - 7.3m) predicting lake depth at broad geographic extents \cite{hollisterPredictingMaximumLake2011, oliver2016prediction, messagerEstimatingVolumeAge2016}. When only a single proxy measure was used, there was a difference in model sensitivity depending on if it was a horizontal distance measure or a vertical slope measure. In the case of a true slope and proxy distance combination, models were more accurate ($\pm$ 4.4m, 29\%) than even the most accurate of prior studies \cite{hollisterPredictingMaximumLake2011, oliver2016prediction, messagerEstimatingVolumeAge2016}. Conversely, models using a proxy slope and true distance combination had prediction error rates ($\pm$ 6.6m, 60\%) of a similar magnitude as that of the baseline proxy-proxy model ($\pm$ 6.4m, 59\%). The greater sensitivity of depth predictions to proxy slope measures relative to proxy distance measures may be explained by the fact that proxy slope measures were a more imperfect representation of true in-lake slopes relative to proxy versus true distances. We did not find evidence that the outsized sensivity of depth predictions to slope was dependent on variations in how these measures were calculated (Table S2). In a general sense, the sensitivity of depth predictions to slope help explain the relatively poor predictive performance of prior non-geometric lake depth models given that they rely heavily on lake area as a predictor \cite{messagerEstimatingVolumeAge2016, oliver2016prediction, sobekPredictingDepthVolume2011} and both horizontal distance measures and vertical slope measures appear to be decoupled from lake area (Figure S9).

### 2020-11-28 manuscript/si_template_2019.tex
text edits

\begin{figure}
      \begin{center}
        \includegraphics[width=\textwidth,keepaspectratio]{../figures/01_slope_compare-1}
      \end{center}
      \caption{Sensitivity analysis regarding different slope calculation methods.}\label{figS1}
    \end{figure}
\clearpage

@@

\begin{figure}
      \noindent\includegraphics[width=0.85\textwidth]{../figures/02_depth_model_importance-1}
      \caption{Importance plot for random forest variables showing increase in mean square error. Higher values indicate greater importance to model predictions. See Equation 1 for a definition of geometric max depth. HUC4 ID is a 'dummy' variable of geographic (hydrologic subbasin) location.}\label{figS5}
      \end{figure}

\begin{figure}
      \begin{center}\includegraphics[width=0.45\textwidth]{../figures/gg_effort-1}\end{center}
            \caption{Comparison between characteristics of lakes with bathymetry data against lakes with depth from other sources in the LAGOSUS-Depth product. The distance to urban area metric is calculated using data from the 2018 US Census Urban and Rural Classification.}\label{figS6}
      \end{figure}

\clearpage
\begin{figure}
      \begin{center}\includegraphics[width=0.85\textwidth]{../figures/01_contrasts-1}\end{center}
      \caption{Comparison of lake characteristics according to differences in lake cross-section shape or reservoir status.}\label{figS7}
      \end{figure}

@@

\begin{figure}
      \begin{center}
        \includegraphics[width=0.79\textwidth,keepaspectratio]{../figures/slope_by_area_depth-1}
      \end{center}
      \caption{Comparison between in-lake and nearshore slopes in concave and convex lakes of the same size and max depth.}\label{figS9}
    \end{figure}
\clearpage

\begin{figure}
      \begin{center}\includegraphics[width=0.73\textwidth]{../figures/lgnemanual-vs-bathy-depth-1}\end{center}
      \caption{Comparison between reported depth and depth extracted from bathymetry surfaces by US State where reported depths come from the LAGOSUS-Depth product. For this figure, no reported depth values originated from the same source as its corresponding bathymetry-derived value.}\label{figS8}
      \end{figure}

>>>>

@@

\begin{figure}
      \begin{center}
        \includegraphics[width=0.79\textwidth,keepaspectratio]{../figures/slope_by_area_depth-1}
      \end{center}
      \caption{Comparison between in-lake and nearshore slopes in concave and convex lakes of the same size and max depth.}\label{figS5}
    \end{figure}
\clearpage

\begin{figure}
      \begin{center}
        \includegraphics[width=\textwidth,keepaspectratio]{../figures/01_slope_compare-1}
      \end{center}
      \caption{Sensitivity analysis regarding different slope calculation methods.}\label{figS6}
    \end{figure}
\clearpage

\begin{figure}
      \noindent\includegraphics[width=0.85\textwidth]{../figures/02_depth_model_importance-1}
      \caption{Importance plot for random forest variables showing increase in mean square error. Higher values indicate greater importance to model predictions. See Equation 1 for a definition of geometric max depth. HUC4 ID is a 'dummy' variable of geographic (hydrologic subbasin) location.}\label{figS7}
      \end{figure}

\begin{figure}
      \begin{center}\includegraphics[width=0.45\textwidth]{../figures/gg_effort-1}\end{center}
            \caption{Comparison between characteristics of lakes with bathymetry data against lakes with depth from other sources in the LAGOSUS-Depth product. The distance to urban area metric is calculated using data from the 2018 US Census Urban and Rural Classification.}\label{figS8}
      \end{figure}

\clearpage
\begin{figure}
      \begin{center}\includegraphics[width=0.85\textwidth]{../figures/01_contrasts-1}\end{center}
      \caption{Comparison of lake characteristics according to differences in lake cross-section shape or reservoir status.}\label{figS9}
      \end{figure}

@@

\begin{figure}
      \begin{center}\includegraphics[width=0.73\textwidth]{../figures/lgnemanual-vs-bathy-depth-1}\end{center}
      \caption{Comparison between reported depth and depth extracted from bathymetry surfaces by US State where reported depths come from the LAGOSUS-Depth product. For this figure, no reported depth values originated from the same source as its corresponding bathymetry-derived value.}\label{figS10}
      \end{figure}
